Temperature and moisture effects on greenhouse gas emissions from deep active-layer boreal soils

### 2016-04-08 manuscript/cpcrw_manuscript.Rmd
Intro and results: new text, summarize lme models

Temperature and moisture typically have strong and often interactive influences on soil GHG emissions. A long literature of laboratory incubations, field observations, and meta-analyses has documented increased CO2 fluxes with rising temperature {Davidson, 2006 #1669}{Hashimoto, 2015 #4231}{Treat, 2015 #4217}. GHG responses to wetting and thawing dynamics are much less certain, with substantial variability between studies {Kim, 2012 #2863}. The anaerobic conditions common following permafrost thaw lower CO~2~ emissions but increase those of CH~4~ {Treat, 2015 #4217;Treat, 2014 #4213}, and interact with soil temperature effects in long-term incubations {Elberling, 2013 #3720}. Such laboratory studies have weaknesses, but their controlled environments provide an important way to elucidate the key mechanisms controlling GHG from high-latitude soils {Schuur, 2015 #4319}. TODO: add Kirschbaum 2006. Most studies have focused on surface soils or permafrost soils, however, neglecting deep soils that experience fluctuating temperature with season and moisture with potential permafrost thaw ..or something less awkward (TODO; CITE). Such deep but unfrozen soils were identified as subject to strong effects from a two-decade warming experiment in the Alaskan Arctic {Sistla, 2013 #3666}.

The goal of this study was to examine how temperature and moisture control C loss and GHG (CO~2~ and CH~4~) emissions from soils sampled from the bottom of the annual active layer--i.e., directly above permafrost--in an Alaskan boreal forest. We also aimed to characterize the chemical and biological properties of these soils as they were subjected to a 100-day incubation at difference temperature and moisture levels. We hypothesized that soils maintained at field moisture and high temperature would lose more C, both as CO~2~ and CH~4~, than cores incubated at 4˚C, due to increased aerobic and anaerobic microbial activity.

@@

For each measurement, a six-core treatment group was connected to a Picarro A0311 multiplexer that was in turn connected to a Picarro G2301 GHG Analyzer (Picarro Inc., Santa Clara, CA, USA). Dry CH~4~ and CO~2~ concentrations were monitored for 2 minutes, and this was repeated 2-3 times before moving on to a new treatment group. Cores were weighed immediately after gas measurements. Ambient air was measured between treatment groups, and before starting measurements in a chamber. BEN: How was ambient air measurements used in this study? don't see any mention of it in remaining text, etc. (TODO)

@@

All data analysis and statistics were performed using R version `r R.version.string` {R Development Core Team, 2016 #4314}. This experiment was run as an "open experiment" with all analysis code, data (from raw instrument data to final summaries), diagnostics, etc., available at https://github.com/bpbond/cpcrw_incubation. The flux data backing the main results have been archived at TODO.

@@

The `r nrow(coredata)` experimental cores had a bulk density of `r pn(mean(bd), 2)` ± `r pn(sd(bd), 2)` (mean ± sd) g cm^-3^. Large (>2 mm) particles, primarily schist, comprised `r round(mean(schist), 0)`% ± `r round(sd(schist), 0)`% of the cores' total mass. Soil (≤2 mm) dry mass was `r round(mean(sdm), 0)` ± `r round(sd(sdm), 0)` g.

Sample DOC was `r pn(cndata_summary["DOC_mg_kg_mean"], 2)` ± `r pn(cndata_summary["DOC_mg_kg_sd"], 2)` mg kg^-1^. Carbon content was `r pn(cndata_summary["C_percent_mean"], 2)`% ± `r pn(cndata_summary["C_percent_sd"], 2)`%, while N content was `r pn(cndata_summary["N_percent_mean"], 2)`% ± `r pn(cndata_summary["N_percent_sd"], 2)`%. Neither temperature was treatment exerted any significant effect on these properties (P > 0.1 for all).

@@

After removal of known problematic data, and exclusion of statistical outliers, CO~2~ flux rates measured during the incubation ranged from `r pn(fluxsummary["flux_µgC_gC_day_min", "CO2"]/1000, 3)` mg C g C^-1^ day^-1^ (`r pn(fluxsummary["flux_µgC_g_day_min", "CO2"]/1000, 3)` mg C g soil^-1^ day^-1^) to a maximum of `r pn(fluxsummary["flux_µgC_gC_day_max", "CO2"]/1000, 3)` (`r pn(fluxsummary["flux_µgC_g_day_max", "CO2"]/1000, 3)`), with a mean of `r pn(fluxsummary["flux_µgC_gC_day_mean", "CO2"]/1000, 3)` (`r pn(fluxsummary["flux_µgC_g_day_mean", "CO2"]/1000, 3)`). CH~4~ rates ranged from `r pn(fluxsummary["flux_µgC_gC_day_min", "CH4"]*1000, 3)` ng C g C^-1^ day^-1^ (`r pn(fluxsummary["flux_µgC_g_day_min", "CH4"]*1000, 3)` ng C g soil^-1^ day^-1^) to a maximum of `r pn(fluxsummary["flux_µgC_gC_day_max", "CH4"]*1000, 3)` (`r pn(fluxsummary["flux_µgC_g_day_max", "CH4"]*1000, 3)`), with a mean of `r pn(fluxsummary["flux_µgC_gC_day_mean", "CH4"]*1000, 3)` (`r pn(fluxsummary["flux_µgC_g_day_mean", "CH4"]*1000, 3)`). Gas fluxes varied considerably over the course of the incubation (**Figures 2 and 3**).

* Statistical tests of WC, T, C, N on individual core fluxes (Table 2)

Cumulative CO~2~ evolved was highly affected by temperature (`r pclean(co2_hsd$Temperature["p adj"])`), and "field moisture" cores emitted significantly more CO~2~ than the other two treatments at both temperatures (P < 0.001 for both). There was no difference between the 20 °C "drought" and "controlled drought" treatments (`r pclean(co2_hsd$Treatment["Drought-Controlled drought", "p adj"])`). Neither temperature (`r pclean(ch4_hsd$Temperature[,"p adj"])`) nor treatment (mean `r pclean(mean(ch4_hsd$Treatment[,"p adj"]))`) was a significant factor in predicting cumulative CH~4~ fluxes.

@@

The production of C from CO~2~ was over `r minmag` order of magnitudes higher than that from CH~4~, with CO~2~:CH~4~ C ratios ranging from `r minratio` million in the `r mintemp` °C "`r maxtrt`" treatment, to `r maxratio` million in the `r maxtemp` °C "`r maxtrt`" treatment.

@@

Morishita et al. {, 2014 #4316} measured GHG fluxes in the field at CPCRW and nearby forests, and found CO~2~ production to be correlated with both soil temperature and moisture, consistent with our results, TODO. Morishita et al. {, 2014 #4316} found however that CH~4~ uptake (they observed no emissions) was driven by temperature only; this is...

Waldrop et al. {, 2010 #4380} incubated  active-layer and permafrost soils from near Fairbanks, AK, under varied temperature and aerobic conditions, observing Q~10~ values of 9.0 (active layer) and 2.3 (permafrost); flux rates were 0.01-0.10 µmol CH~4~ day^-1^ g^-1^ and 1-8 µg C-CO2 hr^-1^ g^-1^.

>>>>

Temperature and moisture typically have strong and often interactive influences on soil GHG emissions. A long literature of laboratory incubations, field observations, and meta-analyses has documented increased CO2 fluxes with rising temperature {Davidson, 2006 #1669}{Hashimoto, 2015 #4231}{Treat, 2015 #4217}. GHG responses to wetting and thawing dynamics are much less certain, with substantial variability between studies {Kim, 2012 #2863}. The anaerobic conditions common following permafrost thaw lower CO~2~ emissions but increase those of CH~4~ {Treat, 2015 #4217;Treat, 2014 #4213}, and interact with soil temperature effects in long-term incubations {Elberling, 2013 #3720}. Such laboratory studies have weaknesses {Podrebarac, 2016 #4365}, but their controlled environments provide an important way to elucidate the key mechanisms controlling GHG from high-latitude soils {Schuur, 2015 #4319}. Most studies have focused on surface soils or permafrost soils, however, neglecting deep active-layer soils that experience fluctuating temperature and, especially, moisture conditions. Such deep but unfrozen soils were identified as subject to strong effects from a two-decade warming experiment in the Alaskan Arctic {Sistla, 2013 #3666}.

The goal of this study was to examine how temperature and moisture control C loss and GHG (CO~2~ and CH~4~) emissions from soils sampled from the bottom of the annual active layer--i.e., directly above permafrost--in an Alaskan boreal forest. We also aimed to characterize the chemical and biological properties of these soils as they were subjected to a 100-day incubation at difference temperature and moisture levels. We hypothesized that (i) CO~2~ would be the dominant pathway for C loss in these largely aerobic soils; (ii) soils maintained at field moisture and high temperature would lose more C-CO~2~ than cores incubated at 4˚C, due to increased aerobic and anaerobic microbial activity; and (iii) core CH~4~ fluxes would be sensitive only to temperature.

@@

For each measurement, a six-core treatment group was connected to a Picarro A0311 multiplexer that was in turn connected to a Picarro G2301 GHG Analyzer (Picarro Inc., Santa Clara, CA, USA). Dry CH~4~ and CO~2~ concentrations were monitored for 2 minutes, and this was repeated 2-3 times before moving on to a new treatment group. Cores were weighed immediately after gas measurements. Ambient air was measured between treatment groups, and before starting measurements in a chamber, as a check on ambient CO~2~ conditions and instrument stability.

@@

All data analysis and statistics were performed using R version `r R.version.string` {R Development Core Team, 2016 #4314}. This experiment was run as an "open experiment" with all analysis code, data (from raw instrument data to final summaries), diagnostics, etc., available at https://github.com/bpbond/cpcrw_incubation. The summarized flux data backing the main results have been archived at [DOI TO BE FILLED IN UPON ACCEPTANCE].

@@

The `r nrow(coredata)` experimental cores had a bulk density of `r pn(mean(bd), 2)` ± `r pn(sd(bd), 2)` (mean ± sd) g cm^-3^. Large (>2 mm) particles, primarily schist, comprised `r round(mean(schist), 0)`% ± `r round(sd(schist), 0)`% of the cores' total mass. Soil (≤2 mm) dry mass was `r round(mean(sdm), 0)` ± `r round(sd(sdm), 0)` g. Sample DOC was `r pn(cndata_summary["DOC_mg_kg_mean"], 2)` ± `r pn(cndata_summary["DOC_mg_kg_sd"], 2)` mg kg^-1^. Carbon content was `r pn(cndata_summary["C_percent_mean"], 2)`% ± `r pn(cndata_summary["C_percent_sd"], 2)`%, while N content was `r pn(cndata_summary["N_percent_mean"], 2)`% ± `r pn(cndata_summary["N_percent_sd"], 2)`%. Neither temperature was treatment exerted any significant effect on these properties (P > 0.1 for all).

@@

After removal of known problematic data, and exclusion of statistical outliers, CO~2~ flux rates measured during the incubation ranged from `r pn(fluxsummary["flux_µgC_gC_day_min", "CO2"], 2)` µg C g C^-1^ day^-1^ (`r pn(fluxsummary["flux_µgC_g_day_min", "CO2"], 2)` µg C g soil^-1^ day^-1^) to a maximum of `r pn(fluxsummary["flux_µgC_gC_day_max", "CO2"], 2)` (`r pn(fluxsummary["flux_µgC_g_day_max", "CO2"], 2)`), with a mean of `r pn(fluxsummary["flux_µgC_gC_day_mean", "CO2"], 3)` (`r pn(fluxsummary["flux_µgC_g_day_mean", "CO2"], 2)`). CH~4~ rates ranged from `r pn(fluxsummary["flux_µgC_gC_day_min", "CH4"]*1000, 2)` ng C g C^-1^ day^-1^ (`r pn(fluxsummary["flux_µgC_g_day_min", "CH4"]*1000, 2)` ng C g soil^-1^ day^-1^) to a maximum of `r pn(fluxsummary["flux_µgC_gC_day_max", "CH4"]*1000, 2)` (`r pn(fluxsummary["flux_µgC_g_day_max", "CH4"]*1000, 3)`), with a mean of `r pn(fluxsummary["flux_µgC_gC_day_mean", "CH4"]*1000, 2)` (`r pn(fluxsummary["flux_µgC_g_day_mean", "CH4"]*1000, 2)`).

Gas fluxes varied considerably over the course of the incubation (**Figures 2 and 3**). In the linear mixed-effects model, CO~2~ was strongly influenced by incubation chamber temperature, core gravimetric water content, and percent soil N (all P < 0.05, and the latter two P < 0.001; **Table 1**). (Note that percent C and percent N were highly correlated (r ≥ 0.98) for these cores. Because percent N was a slightlyl stronger predictor, it was retained in the model while percent C was excluded.) The interaction between water content and soil N was also highly significant (P < 0.001), with high-N cores having little relationship between water content and CO~2~ flux (data not shown). Methane fluxes were most strongly correlated with percent N, while water content exhibited significant interactions with percent N and DOC (**Table 2**). Neither temperature nor water content was a significant first-order predictor of CH~4~ fluxes, however.

The cumulative production of C from CO~2~ was over `r minmag` order of magnitudes higher than that from CH~4~, with CO~2~:CH~4~ C ratios ranging from `r minratio` million in the `r mintemp` °C "`r maxtrt`" treatment, to `r maxratio` million in the `r maxtemp` °C "`r maxtrt`" treatment. Cumulative CO~2~ evolved was highly affected by temperature (`r pclean(co2_hsd$Temperature["p adj"])`), and "field moisture" cores emitted significantly more CO~2~ than the other two treatments at both temperatures (P < 0.001 for both). There was no difference between the 20 °C "drought" and "controlled drought" treatments (`r pclean(co2_hsd$Treatment["Drought-Controlled drought", "p adj"])`). Neither temperature (`r pclean(ch4_hsd$Temperature[,"p adj"])`) nor treatment (mean `r pclean(mean(ch4_hsd$Treatment[,"p adj"]))`) was a significant factor in predicting cumulative CH~4~ fluxes.

@@

@@

Several studies have measured microbial respiration and GHG fluxes from soils very close to our study site. Morishita et al. {, 2014 #4316} measured gas fluxes in the field at CPCRW and nearby forests, and found CO~2~ production to be correlated with both soil temperature and moisture, consistent with our results, TODO. Morishita et al. {, 2014 #4316} found however that CH~4~ uptake (they observed no emissions) was driven by temperature only; this is...

Waldrop et al. {, 2010 #4380} incubated  active-layer and permafrost soils from near Fairbanks, AK, under varied temperature and aerobic conditions, observing Q~10~ values of 9.0 (active layer) and 2.3 (permafrost); flux rates were 0.01-0.10 µmol CH~4~ day^-1^ g^-1^ () and 1-8 µg C-CO2 hr^-1^ g^-1^.

More generally,
